Learning words by recursive pragmatic reasoning about other agents: The blessing of social anxiety

### 2013-05-31 praglearn.tex
edits

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.48\textwidth]{figures/emergence2x2-1.pdf}
\includegraphics[width=0.48\textwidth]{figures/emergence2x2-3.pdf} \\
\includegraphics[width=0.48\textwidth]{figures/emergence3x3-0.pdf}
\includegraphics[width=0.48\textwidth]{figures/emergence3x3-1.pdf} \\
\caption{\label{fig:emergence} Simulations of two learning pragmatic agents playing a naming game. At each stage, $S$ and $L$ have different, possibly contradictory posteriors over the literal lexicon. From these posteriors we derive the probability $P(\text{L understands S})$ (marginalizing over target objects and word choices), and also depict graphically $S$'s model of the listener, $P_{L_n}(\obj | \word, \text{$S$'s data})$, and $L$'s actual model, $P_{L_n}(\obj | \word, \text{$L$'s data})$.}
\end{figure}

@@

What's happening here is that each agent is, effectively, using their partner's behaviour as a basis for forming superstitions about the underlying lexicon that they assume must exist. Since they each then act on these superstitions, and their partner uses the actions to form new superstitions, soon they converge to using similar lexicons, and what started as a superstition becomes normatively correct.

@@

The role of learning here is obvious, but what does the pragmatics part of the model do? It creates the sparse structure in the observed lexicons: at each step, the goal-directed speaker attempts to exploit whatever communicatively useful asymmetries exist in its current model of the listener, and these attempts tend to be successful, since if the listener has even a small asymmetry in their current posterior then their pragmatic reasoning will magnify it as well. But over time, the speaker's use of such patterns, and the listener's ability to understand them, will each be taken by the other as evidence that these patterns actually exist in the literal lexicon. The result is that while both agents start out with uniform priors over the lexicon, they are biased to end up believing in lexicons that are communicatively useful for the task at hand.

@@

A stronger example of how pragmatics can create such biases can be observed if we simulate agents playing this naming game in the ``cheap''/''expensive'' \textsc{common}/\textsc{rare} domain of \S\ref{sec:horn-implicature}. Recall that in this domain, pragmatic reasoning causes each agent to start out independently be biased -- but only weakly -- towards the associations ``cheap'' $\leftrightarrow$ \textsc{common}, ``expensive'' $\leftrightarrow$ \textsc{rare}. Because this bias

>>>>

% You know, this is almost exactly like a quantitatively graded version of Pascal's wager.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.48\textwidth]{figures/emergence2x2-1.pdf}
\includegraphics[width=0.48\textwidth]{figures/emergence2x2-3.pdf} \\
\includegraphics[width=0.48\textwidth]{figures/emergence3x3-0.pdf}
\includegraphics[width=0.48\textwidth]{figures/emergence3x3-1.pdf} \\
\caption{\label{fig:emergence} Simulations of two learning pragmatic agents playing a naming game. At each stage, $S$ and $L$ have different, possibly contradictory posteriors over the literal lexicon. From these posteriors we derive the probability $P(\text{$L$ understands $S$})$ (marginalizing over target objects and word choices), and also depict graphically $S$'s model of the listener, $P_{L_n}(\obj | \word, \text{$S$'s data})$, and $L$'s actual model, $P_{L_n}(\obj | \word, \text{$L$'s data})$.}
\end{figure}

@@

What's happening here is that each agent is, effectively, using their partner's behaviour as a basis for forming superstitions about the underlying lexicon that they assume must exist. Since they then each act on these superstitions, and their partner uses the resulting actions to form new superstitions, they soon converge on using similar lexicons, and what started as a superstition becomes normatively correct.

@@

The role of learning here is obvious, but what does the pragmatics part of the model do? First, it makes it possible for the speaker to learn from listener interpretations, which is not possible without an explicit model of the speaker as a rational agent. Second, it creates the sparse structure in these observed lexicons: at each step, the goal-directed speaker attempts to exploit whatever communicatively useful asymmetries exist in its current model of the listener, and these attempts tend to be successful, since if the listener has even a small asymmetry in their current posterior then their pragmatic reasoning will magnify it as well. But over time, the speaker's use of such patterns, and the listener's ability to understand them, will each be taken by the other as evidence that these patterns actually exist in the literal lexicon. The result is that while both agents start out with uniform priors over the lexicon, they are biased to end up believing in lexicons that are communicatively useful for the task at hand.

@@

A stronger example of how pragmatics can create such biases can be observed if we simulate agents playing this naming game in the ``cheap''/``expensive'' \textsc{common}/\textsc{rare} domain of \S\ref{sec:horn-implicature}. Recall that in this domain, pragmatic reasoning causes each agent to start out independently biased -- but only weakly -- towards the associations ``cheap'' $\leftrightarrow$ \textsc{common}, ``expensive'' $\leftrightarrow$ \textsc{rare}, and that this bias arises exactly because their prior on literal meaning is uniform and uninformative. A fully rational listener, therefore, observing our uncertain speaker using this mapping, would discount it as arising from this bias, and conclude that the speaker in fact did not know which convention was in use. Our socially anxious listener, however, believes that the speaker does know which convention is in use, and therefore tends to take their behaviour as evidence that the good system is in fact in use. Similarly, the socially anxious speaker will wager that since on average they will succed more often if the listener is using the good system, they might as well try it -- and when they succeed, they take this as evidence that the listener was in fact using the good system all along. As a result, dyads in this game not only end up converging onto a stable lexicon, but onto specifically the good lexicon, at a rate far above chance.
